---
title: "Pre-Trends Wald Test with synthdid"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Pre-Trends Wald Test with synthdid}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 7,
  fig.height = 4
)
```

## Overview

This vignette demonstrates the full `synthdid` estimation workflow using the
`california_prop99` dataset, then introduces `calculate_wald_pre_test()` — a
joint Wald test of the no-pre-trends assumption.

The vignette is structured in two parts:

1. **Standard workflow** — point estimate, standard error, confidence interval,
   and a diagnostic plot using `california_prop99`.
2. **Pre-trends Wald test** — explanation of why a single-cohort dataset cannot
   exercise the test, followed by a simulation of staggered adoption data that
   can.

---

## Part 1: Standard Estimation Workflow

### Data

`california_prop99` records annual per-capita cigarette consumption (packs) for
39 US states from 1970–2000. California imposed a tobacco tax in 1989, making
it the single treated unit.

```{r load-data}
library(synthdid)

data("california_prop99")
head(california_prop99)
```

### Convert to matrix form

`panel.matrices()` reshapes the long panel into the N × T outcome matrix `Y`
required by the estimators, and returns `N0` (number of control units) and `T0`
(number of pre-treatment periods).

```{r panel-matrices}
setup <- panel.matrices(california_prop99)

cat("Units (N)       :", nrow(setup$Y), "\n")
cat("Control units   :", setup$N0, "\n")
cat("Time periods (T):", ncol(setup$Y), "\n")
cat("Pre-treatment   :", setup$T0, "\n")
```

### Estimate the treatment effect

```{r estimate}
tau.hat <- synthdid_estimate(setup$Y, setup$N0, setup$T0)
tau.hat
```

### Standard error and confidence interval

We use the placebo variance estimator (the only option available when there is
a single treated unit).

```{r se-ci}
vcov_mat <- vcov(tau.hat, method = "placebo")
se       <- sqrt(vcov_mat)

sprintf("Point estimate : %1.2f", tau.hat)
sprintf("95%% CI         : (%1.2f, %1.2f)",
        tau.hat - 1.96 * se,
        tau.hat + 1.96 * se)
```

The estimate suggests that Proposition 99 reduced cigarette consumption by
roughly `r round(abs(c(tau.hat)), 1)` packs per capita per year.

### Diagnostic plot

```{r plot, fig.alt="synthdid estimate plot for California Prop 99"}
plot(tau.hat)
```

---

## Part 2: Pre-Trends Wald Test

### Why `california_prop99` cannot be used here

`calculate_wald_pre_test()` tests the null hypothesis that all pre-treatment
ATT(g, t) estimates are jointly zero — i.e. the parallel-trends assumption
holds before treatment begins. This requires **multiple (cohort, calendar-time)
pairs** where `t < g`.

`california_prop99` has only one treated cohort (California, `g = 1989`) and
one treatment period. There are therefore no pre-treatment ATT(g, t) cells at
all: the index set `{(g, t) : t < g}` is empty. Running the test on this
dataset would return immediately with the message *"No pre-treatment periods
found to test."*

The function is designed for **staggered adoption** settings — multiple cohorts
each treated at different times — as studied in Callaway & Sant'Anna (2021).

### Simulating staggered adoption data

We simulate a balanced panel with three treatment cohorts to produce a dataset
where the Wald test is meaningful.

```{r simulate}
set.seed(42)

# Panel dimensions
N_ctrl <- 50   # control units
T_total <- 10  # total time periods

# Three cohorts treated at periods 4, 6, and 8 respectively
cohort_specs <- list(
  list(g = 4, n = 10),
  list(g = 6, n = 10),
  list(g = 8, n = 10)
)
N_treated <- sum(sapply(cohort_specs, `[[`, "n"))
N_total   <- N_ctrl + N_treated

# True treatment effect (zero in pre-period by construction)
tau_true <- 2

# Generate outcome matrix: unit fixed effect + time trend + noise
unit_fe <- rnorm(N_total, mean = 0, sd = 1)
time_fe <- seq(0, 1, length.out = T_total)

Y <- outer(unit_fe, rep(1, T_total)) +
     outer(rep(1, N_total), time_fe) +
     matrix(rnorm(N_total * T_total, sd = 0.5), N_total, T_total)

# Build (cohort, time, ATT, SE) structure manually to mimic
# what a DiD package would return for all (g, t) pairs
cohort_g <- c()
period_t <- c()

row <- N_ctrl + 1
for (spec in cohort_specs) {
  g  <- spec$g
  n1 <- spec$n

  # Add treatment effect in post-period rows
  post_cols <- g:T_total
  Y[row:(row + n1 - 1), post_cols] <-
    Y[row:(row + n1 - 1), post_cols] + tau_true

  # Record all (g, t) pairs for this cohort
  cohort_g <- c(cohort_g, rep(g, T_total))
  period_t <- c(period_t, 1:T_total)

  row <- row + n1
}

cat("(cohort, time) pairs recorded:", length(cohort_g), "\n")
cat("Pre-treatment pairs (t < g)  :",
    sum(cohort_g > period_t), "\n")
```

### Simulate ATT estimates and a covariance matrix

In practice these come from a DiD estimator (e.g. `att_gt()` in the `did`
package). Here we generate plausible values directly.

```{r att-vcov}
n_pairs <- length(cohort_g)

# Simulate ATT(g, t) estimates:
#   - Pre-treatment (t < g):  mean 0, small noise  => should pass the test
#   - Post-treatment (t >= g): mean tau_true
is_pre  <- cohort_g > period_t
att_est <- ifelse(is_pre,
                  rnorm(n_pairs, mean = 0,        sd = 0.15),
                  rnorm(n_pairs, mean = tau_true,  sd = 0.20))

# Diagonal covariance with small, realistic variances
se_vals  <- ifelse(is_pre, 0.15, 0.20)
vcov_mat_sim <- diag(se_vals^2)
```

### Run the pre-trends Wald test

`vcov_is_normalized = TRUE` tells the function that `vcov_mat_sim` is the
**normalized** asymptotic covariance Cov(att), so the statistic is multiplied
by `n_scaling_factor` to reach the chi-squared scale. Set
`vcov_is_normalized = FALSE` if your DiD package returns an already n-scaled
matrix.

```{r wald-test}
result <- calculate_wald_pre_test(
  att_estimates         = att_est,
  cohort_treatment_time = cohort_g,
  calendar_time_period  = period_t,
  vcov_matrix_all_atts  = vcov_mat_sim,
  std_errors_all_atts   = se_vals,
  n_scaling_factor      = N_total,
  vcov_is_normalized    = TRUE,
  verbose               = TRUE
)
```

```{r wald-results}
cat("Wald statistic :", round(result$W,     4), "\n")
cat("Degrees of freedom:", result$q,            "\n")
cat("p-value        :", round(result$Wpval, 6), "\n")
```

A large p-value (typically > 0.05) indicates no evidence against the
no-pre-trends null — the parallel-trends assumption is supported in the
pre-treatment periods.

### Testing a linear contrast

Rather than testing every pre-treatment period individually, you may want to
test a specific restriction — for example, that the average pre-treatment
effect across all cohort-period pairs is zero. Supply a single-row contrast
matrix `R`:

```{r wald-contrast}
n_pre <- sum(cohort_g > period_t)
R_avg <- matrix(rep(1 / n_pre, n_pre), nrow = 1)

result_avg <- calculate_wald_pre_test(
  att_estimates         = att_est,
  cohort_treatment_time = cohort_g,
  calendar_time_period  = period_t,
  vcov_matrix_all_atts  = vcov_mat_sim,
  std_errors_all_atts   = se_vals,
  n_scaling_factor      = N_total,
  vcov_is_normalized    = TRUE,
  R                     = R_avg,
  verbose               = FALSE
)

sprintf("Average pre-trend test: W = %.4f, df = %d, p-value = %.4f",
        result_avg$W, result_avg$q, result_avg$Wpval)
```

---

## Summary

| Step | Function |
|---|---|
| Reshape panel data | `panel.matrices()` |
| Estimate ATT | `synthdid_estimate()`, `sc_estimate()`, `did_estimate()` |
| Variance / SE | `vcov()`, `synthdid_se()` |
| Confidence interval | Manual: `tau.hat ± 1.96 * se` |
| Diagnostic plot | `plot()` |
| Pre-trends Wald test | `calculate_wald_pre_test()` |

### Key argument: `vcov_is_normalized`

| Your covariance matrix | Correct setting |
|---|---|
| `Cov(att)` — normalized (e.g. from most DiD packages) | `vcov_is_normalized = TRUE` + supply `n_scaling_factor` |
| `n * Cov(att)` — already n-scaled (e.g. internal C&S format) | `vcov_is_normalized = FALSE` |

Passing the wrong convention inflates or deflates the Wald statistic by a
factor of n, so it is important to check how your DiD estimator scales its
output covariance matrix.

---

## References

- Arkhangelsky, D., Athey, S., Hirshberg, D.A., Imbens, G.W., and Wager, S.
  (2021). "Synthetic Difference in Differences." *American Economic Review*,
  111(12), 4088–4118.

- Callaway, B. and Sant'Anna, P.H.C. (2021). "Difference-in-Differences with
  Multiple Time Periods." *Journal of Econometrics*, 225(2), 200–230.

- Abadie, A., Diamond, A., and Hainmueller, J. (2010). "Synthetic Control
  Methods for Comparative Case Studies." *Journal of the American Statistical
  Association*, 105(490), 493–505.
